---
title: "Seed Search Results"
subtitle: "Initial HMM search statistics for eukaryotic 2A peptides"
date: today
execute:
  echo: true
  warning: false
---

## Overview

This page summarizes the initial HMM search results using seed models built from curated 2A peptide alignments.

## Seed Alignments

The seed alignments contain known, experimentally validated 2A peptides:

| Class | Peptides | Source |
|-------|----------|--------|
| Class 1 | T2A, E2A, P2A | Picornaviruses |
| Class 2 | F2A variants | Aphthoviruses |

## Search Statistics

```{r}
#| label: tbl-search-stats
#| tbl-cap: "HMM search statistics by database"
#| eval: false

library(tidyverse)

# Load search statistics (when available)
results_dir <- Sys.getenv("RESULTS_DIR", "../../scratch/results")

# Example statistics table
stats <- tibble(
    Database = c('UniProt', 'RefProt', 'UniParc', 'MGnify'),
    `Class 1 Hits` = c(45, 128, 892, 3421),
    `Class 2 Hits` = c(23, 67, 445, 1893),
    `E-value < 1e-10` = c(68, 195, 1337, 5314)
)

stats
```

## Hit Distribution

### By Database

::: {.callout-note}
## Placeholder
Actual search results will be displayed here once the pipeline has been run.
:::

### By E-value

```{r}
#| label: fig-evalue-dist
#| fig-cap: "Distribution of E-values from seed HMM searches"
#| fig-width: 8
#| fig-height: 5
#| eval: false

library(ggplot2)

# Example E-value distribution
evalues <- rexp(1000, rate = 1e8)
df <- tibble(evalue = evalues)

ggplot(df, aes(x = evalue)) +
    geom_histogram(bins = 50) +
    scale_x_log10() +
    labs(x = "E-value", y = "Count") +
    theme_minimal()
```

## Quality Metrics

### Alignment Length Distribution

Expected alignment lengths:
- Class 1: 18-22 residues
- Class 2: 20-24 residues

### Conservation Scores

Key conserved positions in seed alignments:
- Position -1 (G): >95% conserved
- Position 0 (P): 100% conserved
- Position -4 (N/P): >90% conserved

## Filtering Summary

| Filter | Sequences Removed |
|--------|-------------------|
| E-value > 1e-5 | TBD |
| Length < 15 | TBD |
| Gap % > 50% | TBD |
| Missing PGP motif | TBD |

## Next Steps

1. Review filtered alignments
2. Build refined models from high-confidence hits
3. Perform iteration 2 searches
4. Generate sequence logos
