---
title: "Sequence Logos"
subtitle: "Visualization of conserved motifs in 2A peptides"
date: today
---

```{r}
#| label: setup
#| include: false

library(ggseqlogo)
library(ggplot2)

# Helper function to read gzipped Stockholm alignment and extract sequences
read_stockholm_sequences <- function(sto_path) {
  # Read gzipped Stockholm file
  lines <- readLines(gzfile(sto_path))

  # Filter to sequence lines (not comments, not // delimiter, not empty)
  seq_lines <- lines[!grepl("^#", lines) &
                     !grepl("^//", lines) &
                     nchar(trimws(lines)) > 0]

  # Parse sequence lines: "name  sequence"
  sequences <- sapply(seq_lines, function(line) {
    parts <- strsplit(trimws(line), "\\s+")[[1]]
    # Return the sequence part (second element), converting . to - for gaps
    gsub("\\.", "-", parts[2])
  }, USE.NAMES = FALSE)

  return(sequences)
}

# Paths to seed alignments
class1_path <- "../../resources/seed-alignments/2A-class-1.sto.gz"
class2_path <- "../../resources/seed-alignments/2A-class-2.sto.gz"
```

## Overview

Sequence logos visualize the conservation pattern at each position in aligned 2A peptides. Height indicates information content (conservation), and letter size shows relative amino acid frequency.

## Class 1 Sequence Logo

```{r}
#| label: fig-class1-logo
#| fig-cap: "Class 1 2A peptide sequence logo showing conserved GDVE central motif and NPGP C-terminal motif"
#| fig-width: 10
#| fig-height: 3

class1_seqs <- read_stockholm_sequences(class1_path)
message(sprintf("Loaded %d Class 1 sequences", length(class1_seqs)))

ggseqlogo(class1_seqs, method = "prob") +
  labs(
    title = "Class 1 2A Peptide",
    x = "Position",
    y = "Probability"
  ) +
  theme_logo() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 10)
  )
```

### Expected Conservation Pattern

```
Position:  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18
Consensus: L  L  x  x  x  x  G  D  V  E  x  x  x  x  N  P  G  P
                          ^^^^^^^               ^^^^^^^^^^
                          Central               C-terminal
```

**Key positions:**
- Positions 7-10: GDVE central motif
- Positions 15-18: NPGP C-terminal motif

## Class 2 Sequence Logo

```{r}
#| label: fig-class2-logo
#| fig-cap: "Class 2 2A peptide sequence logo showing conserved N-terminal tryptophan, EEGIE central motif, and PNPGP C-terminal motif"
#| fig-width: 10
#| fig-height: 3

class2_seqs <- read_stockholm_sequences(class2_path)
message(sprintf("Loaded %d Class 2 sequences", length(class2_seqs)))

ggseqlogo(class2_seqs, method = "prob") +
  labs(
    title = "Class 2 2A Peptide",
    x = "Position",
    y = "Probability"
  ) +
  theme_logo() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 10)
  )
```

### Expected Conservation Pattern

```
Position:  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19
Consensus: W  x  x  x  x  E  E  G  I  E  x  x  x  P  N  P  G  P  x
           ^              ^^^^^^^^^^^        ^^^^^^^^^^^^^^
           N-terminal     Central            C-terminal
```

**Key positions:**
- Position 1: Invariant tryptophan (W)
- Positions 6-10: EEGIE central motif
- Positions 14-18: PNPGP C-terminal motif

## Interpretation Guide

| Feature | Meaning |
|---------|---------|
| Tall letters | Highly conserved position |
| Short letters | Variable position |
| Stack height | Information content (bits) |
| Letter order | Frequency (top = most common) |

## Conservation by Iteration

Track how conservation patterns change through iterative refinement:

| Iteration | Sequences | Avg. Conservation |
|-----------|-----------|-------------------|
| Seed | ~10 | High |
| Iter 1 | ~100 | Medium |
| Iter 2 | ~500 | Lower (more diversity) |
| Final (curated) | ~200 | Medium-High |

## Position-Specific Features

### Critical Residues

| Position | Residue | Function | Tolerance |
|----------|---------|----------|-----------|
| -1 | G | Cleavage site | None |
| 0 | P | Cleavage site | None |
| -2 | G | Positioning | Low |
| -3 | P/N | Efficiency | Medium |

### Variable Regions

- Positions 2-6: Allow diversity
- Positions 11-14: Linker region
- Tolerate substitutions without losing function
