---
title: "2A Peptide Domain Co-occurrence Analysis"
author: "2A Peptide Search Pipeline"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
params:
  class: "class-1"
  interproscan_file: null
  domains_file: null
  results_dir: "scratch/results"
editor: visual
editor_options:
  chunk_output_type: console
---

## Domain Analysis for 2A-`r params$class`

This report analyzes the domain organization of proteins containing 2A peptides,
examining which protein domains commonly flank 2A peptide sequences.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(janitor)
library(DT)
library(cowplot)
library(glue)

# Set up paths based on parameters
results_dir <- params$results_dir
peptide_class <- params$class

# Construct file paths
interproscan_file <- params$interproscan_file
if (is.null(interproscan_file)) {
  interproscan_file <- file.path(results_dir, "domains", paste0("interproscan.", peptide_class, ".tsv.gz"))
}

domains_file <- params$domains_file
if (is.null(domains_file)) {
  domains_file <- file.path(results_dir, "domains", paste0("2a_domains.", peptide_class, ".tsv"))
}
```

## Data Loading

```{r}
#| label: load_data
#| message: false

# InterProScan column names
ipr_colnames <- c(
  'protein_acc', 'seq_md5', 'seq_len', 'sig_db', 'sig_acc',
  'sig_desc', 'start_loc', 'stop_loc', 'score', 'status',
  'date', 'ipr_annot_acc', 'ipr_annot_desc'
)

# Load InterProScan results if available
has_interproscan <- file.exists(interproscan_file)

if (has_interproscan) {
  iprscan_tbl <- read_tsv(
    interproscan_file,
    col_names = ipr_colnames,
    show_col_types = FALSE
  ) |>
    select(-c(seq_md5, status, date)) |>
    rename(
      dom_evalue = score,
      query_start = start_loc,
      query_end = stop_loc
    ) |>
    mutate(dom_evalue = as.numeric(dom_evalue)) |>
    mutate(
      seq_db = case_when(
        str_detect(protein_acc, "^tr") ~ "uniprot",
        str_detect(protein_acc, "^sp") ~ "uniprot",
        str_detect(protein_acc, "^MGY") ~ "mgnify",
        str_detect(protein_acc, "^UPI") ~ "uniparc",
        .default = "unknown"
      )
    ) |>
    mutate(
      seq_db = as_factor(seq_db),
      sig_db = as_factor(sig_db)
    )

  message(glue("Loaded {nrow(iprscan_tbl)} InterProScan annotations"))
} else {
  message("InterProScan results not available - skipping external domain analysis")
  iprscan_tbl <- tibble()
}

# Load 2A domain hits
twoa_domains_tbl <- read_tsv(
  domains_file,
  show_col_types = FALSE
) |>
  janitor::clean_names() |>
  filter(dom_evalue <= 0.01) |>
  mutate(
    seq_db = case_when(
      str_detect(protein_acc, "^tr") ~ "uniprot",
      str_detect(protein_acc, "^sp") ~ "uniprot",
      str_detect(protein_acc, "^MGY") ~ "mgnify",
      str_detect(protein_acc, "^UPI") ~ "uniparc",
      .default = "unknown"
    )
  ) |>
  mutate(
    seq_db = as_factor(seq_db),
    sig_db = as_factor(sig_db)
  )

message(glue("Loaded {nrow(twoa_domains_tbl)} 2A domain hits"))
```

```{r}
#| label: combine_domains
#| eval: !expr has_interproscan

# Combine all domain annotations
if (has_interproscan) {
  # Extract InterPro accessions as separate entries
  ipr_db_tbl <- iprscan_tbl |>
    select(
      protein_acc, sig_acc = ipr_annot_acc, sig_desc = ipr_annot_desc,
      query_start, query_end, dom_evalue, seq_db
    ) |>
    filter(sig_acc != '-', !is.na(sig_acc)) |>
    mutate(sig_db = 'InterPro')

  domains_tbl <- bind_rows(
    iprscan_tbl |> select(protein_acc, sig_db, sig_acc, sig_desc, query_start, query_end, dom_evalue, seq_db),
    twoa_domains_tbl |> select(protein_acc, sig_db, sig_acc, sig_desc, query_start, query_end, dom_evalue, seq_db),
    ipr_db_tbl
  ) |>
    arrange(protein_acc, query_start) |>
    unique()
} else {
  domains_tbl <- twoa_domains_tbl |>
    select(protein_acc, sig_db, sig_acc, sig_desc, query_start, query_end, dom_evalue, seq_db) |>
    arrange(protein_acc, query_start)
}

# Clean up UniProt IDs (remove tr|XXX| prefix)
domains_tbl <- domains_tbl |>
  mutate(
    protein_acc = case_when(
      str_detect(protein_acc, "^(tr|sp)\\|") ~ str_extract(protein_acc, "(?<=\\|)[^|]+(?=\\|)"),
      .default = protein_acc
    )
  )
```

## Domain Counts per Protein

```{r}
#| label: domain_counts
domain_cnts <- domains_tbl |>
  filter(sig_db %in% c('InterPro', 'Hmmer3-custom', 'Pfam') | is.na(dom_evalue)) |>
  count(protein_acc, seq_db, sort = TRUE, name = 'n_dom')

protein_domain_cnts <- domain_cnts |>
  mutate(n_dom = as_factor(n_dom)) |>
  count(n_dom)

ggplot(protein_domain_cnts, aes(x = n_dom, y = n)) +
  geom_col(fill = 'steelblue') +
  theme_cowplot() +
  labs(
    x = "Number of annotated domains",
    y = "Number of proteins",
    title = glue("Domain annotations per protein (2A-{peptide_class})"),
    subtitle = glue("Total proteins = {scales::comma(nrow(domain_cnts))}")
  )
```

## Domain Neighbors

The following analysis identifies which domains commonly flank 2A peptides on the
N-terminal and C-terminal sides.

```{r}
#| label: domain_neighbors
domains_filt_tbl <- domains_tbl |>
  filter(sig_db %in% c('InterPro', 'Hmmer3-custom', 'Pfam') | is.na(dom_evalue))

domain_org <- domains_filt_tbl |>
  arrange(protein_acc, query_start) |>
  mutate(
    nterm = lag(sig_acc),
    nterm_dist = query_start - lag(query_end),
    cterm = lead(sig_acc),
    cterm_dist = lead(query_start) - query_end,
    .by = protein_acc
  )

twoa_neighbors <- domain_org |>
  filter(str_detect(sig_acc, "2A-class")) |>
  select(protein_acc, seq_db, nterm, sig_acc, cterm)

neighbor_counts <- twoa_neighbors |>
  count(seq_db, nterm, sig_acc, cterm, sort = TRUE)

DT::datatable(
  neighbor_counts,
  caption = glue("Domain neighbors for 2A-{peptide_class} peptides"),
  filter = 'top',
  options = list(pageLength = 25)
)
```

## Most Common N-terminal Domains

```{r}
#| label: nterm_domains
nterm_counts <- twoa_neighbors |>
  filter(!is.na(nterm)) |>
  count(nterm, sort = TRUE) |>
  head(20)

ggplot(nterm_counts, aes(x = reorder(nterm, n), y = n)) +
  geom_col(fill = 'darkgreen') +
  coord_flip() +
  theme_cowplot() +
  labs(
    x = "N-terminal domain",
    y = "Count",
    title = glue("Top 20 N-terminal domains (2A-{peptide_class})")
  )
```

## Most Common C-terminal Domains

```{r}
#| label: cterm_domains
cterm_counts <- twoa_neighbors |>
  filter(!is.na(cterm)) |>
  count(cterm, sort = TRUE) |>
  head(20)

ggplot(cterm_counts, aes(x = reorder(cterm, n), y = n)) +
  geom_col(fill = 'darkred') +
  coord_flip() +
  theme_cowplot() +
  labs(
    x = "C-terminal domain",
    y = "Count",
    title = glue("Top 20 C-terminal domains (2A-{peptide_class})")
  )
```

## Database Distribution

```{r}
#| label: db_distribution
db_counts <- twoa_neighbors |>
  count(seq_db, sort = TRUE)

ggplot(db_counts, aes(x = reorder(seq_db, n), y = n, fill = seq_db)) +
  geom_col() +
  coord_flip() +
  theme_cowplot() +
  theme(legend.position = 'none') +
  labs(
    x = "Database",
    y = "Number of 2A peptides",
    title = glue("2A-{peptide_class} hits by database")
  )
```

## Session Info

```{r}
#| label: session_info
sessionInfo()
```
