---
title: "2A Peptide Model Validation Report"
author: "2A Peptide Search Pipeline"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
params:
  results_dir: "scratch/results"
editor: visual
editor_options:
  chunk_output_type: console
---

## Model Validation Against Literature

This report evaluates the sensitivity of the trained 2A peptide HMM models
against a curated set of published 2A peptide sequences with known activity
measurements.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
library(cowplot)
library(glue)
library(scales)

results_dir <- params$results_dir
comparison_file <- file.path(results_dir, "validation", "literature_comparison.tsv")
```

## Load Comparison Data

```{r}
#| label: load_data
comparison_tbl <- read_tsv(comparison_file, show_col_types = FALSE) |>
  mutate(
    activity_numeric = case_when(
      activity == "Yes" ~ 100,
      activity == "N/A" ~ NA_real_,
      str_detect(activity, "%") ~ as.numeric(str_remove(activity, "%")),
      .default = NA_real_
    ),
    class_original = case_when(
      class_original == "A" ~ "Class 1 (A)",
      class_original == "B" ~ "Class 2 (B)",
      .default = class_original
    )
  )

n_total <- nrow(comparison_tbl)
n_hits <- sum(comparison_tbl$classification != "no-hit")
```

## Overall Detection Summary

Of **`r scales::comma(n_total)`** literature 2A sequences, the models detected
**`r scales::comma(n_hits)`** (**`r round(100*n_hits/n_total, 1)`%**).

```{r}
#| label: detection_summary
classification_counts <- comparison_tbl |>
  count(classification, sort = TRUE)

ggplot(classification_counts, aes(x = reorder(classification, n), y = n, fill = classification)) +
  geom_col() +
  coord_flip() +
  theme_cowplot() +
  theme(legend.position = "none") +
  labs(
    x = "Classification",
    y = "Number of sequences",
    title = "HMM Model Detection Results"
  ) +
  scale_fill_brewer(palette = "Set2")
```

## Detection by Original Class

Comparison of HMM classification against the original class annotations from
the literature.

```{r}
#| label: class_comparison
class_comparison <- comparison_tbl |>
  mutate(
    hmm_class = case_when(
      str_detect(classification, "class-1") ~ "HMM Class 1",
      str_detect(classification, "class-2") ~ "HMM Class 2",
      .default = "Not Detected"
    )
  ) |>
  count(class_original, hmm_class) |>
  pivot_wider(names_from = hmm_class, values_from = n, values_fill = 0)

knitr::kable(class_comparison, caption = "Literature class vs HMM classification")
```

## Detection by Source

```{r}
#| label: detection_by_source
source_detection <- comparison_tbl |>
  mutate(detected = classification != "no-hit") |>
  group_by(source) |>
  summarize(
    total = n(),
    detected = sum(detected),
    pct_detected = 100 * detected / total,
    .groups = "drop"
  ) |>
  arrange(desc(pct_detected))

ggplot(source_detection, aes(x = reorder(source, pct_detected), y = pct_detected, fill = source)) +
  geom_col() +
  geom_text(aes(label = glue("{detected}/{total}")), hjust = -0.1, size = 3) +
  coord_flip() +
  ylim(0, 110) +
  theme_cowplot() +
  theme(legend.position = "none") +
  labs(
    x = "Source",
    y = "Detection rate (%)",
    title = "Detection Rate by Source Type"
  )
```

## Activity Analysis

For sequences with measured activity, we can examine whether the HMM models
preferentially detect sequences with higher 2A activity.

```{r}
#| label: activity_analysis
activity_tbl <- comparison_tbl |>
  filter(!is.na(activity_numeric)) |>
  mutate(detected = classification != "no-hit")

if (nrow(activity_tbl) > 10) {
  ggplot(activity_tbl, aes(x = detected, y = activity_numeric, fill = detected)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    theme_cowplot() +
    theme(legend.position = "none") +
    labs(
      x = "Detected by HMM",
      y = "Activity (%)",
      title = "Activity Distribution by Detection Status"
    ) +
    scale_fill_brewer(palette = "Set1")
} else {
  message("Insufficient activity data for analysis")
}
```

## Missed Sequences

Sequences that were not detected by either model may represent:
- Novel 2A variants outside the current model scope
- Sequences with very low activity (false positives in literature)
- Edge cases requiring model refinement

```{r}
#| label: missed_sequences
missed_tbl <- comparison_tbl |>
  filter(classification == "no-hit") |>
  select(name, sequence, source, class_original, activity)

if (nrow(missed_tbl) > 0) {
  DT::datatable(
    missed_tbl,
    caption = "Sequences not detected by HMM models",
    filter = 'top',
    options = list(pageLength = 25, scrollX = TRUE)
  )
} else {
  message("All sequences were detected!")
}
```

## Detailed Results Table

```{r}
#| label: full_table
DT::datatable(
  comparison_tbl |>
    select(name, sequence, source, class_original, activity,
           hmm_class1_hit, hmm_class1_score,
           hmm_class2_hit, hmm_class2_score,
           classification),
  caption = "Complete validation results",
  filter = 'top',
  options = list(pageLength = 25, scrollX = TRUE)
)
```

## Session Info

```{r}
#| label: session_info
sessionInfo()
```
