---
title: "Prokaryotic Ribosomal Stalling Peptide Discovery"
subtitle: "GP Motif Analysis and Pattern Discovery"
author: "2A Peptide Search Pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)
library(here)

# Snakemake inputs
comparison_file <- snakemake@input$comparison
validation_file <- snakemake@input$validation
conservation_file <- snakemake@input$conservation
consensus_file <- snakemake@input$consensus

# Output
output_file <- snakemake@output$report
```

# Executive Summary

This report presents the results of a comprehensive search for prokaryotic ribosomal stalling peptides using two complementary approaches:

1. **Seed-based discovery**: Starting from known stalling peptides (SecM, TnaC, MifM, etc.)
2. **Unbiased GP discovery**: Extracting all GP-containing sequences and focusing on inter-domain regions

```{r load-data}
# Load comparison statistics
comparison <- read_tsv(comparison_file, show_col_types = FALSE)

# Load validation results
if (file.exists(validation_file) && file.size(validation_file) > 0) {
  validation <- read_tsv(validation_file, show_col_types = FALSE)
} else {
  validation <- tibble()
}

# Load conservation data
conservation <- read_tsv(conservation_file, show_col_types = FALSE)

# Load consensus patterns
consensus <- read_tsv(consensus_file, show_col_types = FALSE)
```

## Key Findings

**Discovery Statistics:**

- **Total GP motifs found**: `r format(comparison$total_gp_motifs, big.mark=",")`
- **Inter-domain GP motifs**: `r format(comparison$interdomain_gp_motifs, big.mark=",")` (`r sprintf("%.1f%%", comparison$interdomain_fraction * 100)`)
- **Clusters identified**: `r comparison$n_clusters`
- **Known peptides recovered**: `r comparison$known_peptides_recovered` (validation)

**Conservation:**

- **High-quality patterns**: `r nrow(consensus)` consensus patterns identified
- **Mean conservation**: `r sprintf("%.3f", mean(conservation$mean_conservation))`
- **Conservation range**: `r sprintf("%.3f", min(conservation$mean_conservation))` - `r sprintf("%.3f", max(conservation$mean_conservation))`

---

# Approach Comparison

## GP Motif Distribution

### Overview Statistics

```{r stats-table}
stats_summary <- tibble(
  Metric = c(
    "Total GP motifs",
    "Inter-domain GP",
    "Intra-domain GP",
    "Proteins with GP",
    "Proteins with inter-domain GP",
    "Inter-domain enrichment"
  ),
  Value = c(
    format(comparison$total_gp_motifs, big.mark=","),
    format(comparison$interdomain_gp_motifs, big.mark=","),
    format(comparison$intradomain_gp_motifs, big.mark=","),
    format(comparison$proteins_with_gp, big.mark=","),
    format(comparison$proteins_with_interdomain_gp, big.mark=","),
    sprintf("%.1f%%", comparison$interdomain_fraction * 100)
  )
)

kable(stats_summary, caption = "GP Motif Discovery Statistics")
```

### Interpretation

The analysis identified **`r format(comparison$total_gp_motifs, big.mark=",")`** GP motifs across the prokaryotic proteome. Of these, **`r sprintf("%.1f%%", comparison$interdomain_fraction * 100)`** occur in inter-domain regions (between protein domains).

This enrichment of GP motifs in inter-domain regions supports the hypothesis that ribosomal stalling peptides preferentially occur at domain boundaries, where:

1. Structural transitions may facilitate ribosomal pausing
2. Co-translational domain folding creates steric constraints
3. Regulatory control of multi-domain protein assembly is advantageous

## Comparison Plots

```{r comparison-plots, fig.height=8}
# Display comparison plots generated by compare_gp_approaches.py
plots_dir <- dirname(comparison_file)
plots_dir <- file.path(dirname(plots_dir), "comparison_plots")

if (dir.exists(plots_dir)) {
  summary_plot <- file.path(plots_dir, "summary_comparison.png")
  if (file.exists(summary_plot)) {
    knitr::include_graphics(summary_plot)
  }
}
```

---

# Pattern Discovery

## Cluster Analysis

```{r cluster-stats}
cluster_stats <- tibble(
  Metric = c(
    "Total clusters",
    "Mean cluster size",
    "Median cluster size",
    "Largest cluster",
    "Smallest cluster"
  ),
  Value = c(
    comparison$n_clusters,
    sprintf("%.1f", comparison$mean_cluster_size),
    sprintf("%.0f", comparison$median_cluster_size),
    comparison$max_cluster_size,
    comparison$min_cluster_size
  )
)

kable(cluster_stats, caption = "Clustering Statistics")
```

## Conservation Patterns

### Top Conserved Clusters

```{r top-conserved}
top_conserved <- conservation %>%
  arrange(desc(mean_conservation)) %>%
  head(10) %>%
  select(cluster_id, cluster_size, mean_conservation, consensus_sequence, gp_position)

kable(top_conserved,
      col.names = c("Cluster", "Size", "Conservation", "Consensus Sequence", "GP Position"),
      caption = "Top 10 Most Conserved Clusters",
      digits = 3)
```

### Cluster Size vs Conservation

```{r conservation-plot}
ggplot(conservation, aes(x = cluster_size, y = mean_conservation)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  scale_x_log10() +
  labs(
    title = "Cluster Size vs Conservation",
    x = "Cluster Size (log scale)",
    y = "Mean Conservation Score",
    caption = "Each point represents one cluster"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12)
  )
```

---

# Consensus Patterns

## Motif Classification

```{r motif-types}
motif_summary <- consensus %>%
  count(motif_type, name = "count") %>%
  arrange(desc(count))

ggplot(motif_summary, aes(x = reorder(motif_type, count), y = count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Distribution of Motif Types",
    x = "Motif Type",
    y = "Number of Clusters"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )
```

## High-Quality Consensus Patterns

```{r high-quality-patterns}
high_quality <- consensus %>%
  arrange(desc(quality_score)) %>%
  head(15) %>%
  select(cluster_id, motif_type, cluster_size, mean_conservation,
         quality_score, consensus_sequence)

kable(high_quality,
      col.names = c("Cluster", "Motif Type", "Size", "Conservation",
                    "Quality", "Consensus Sequence"),
      caption = "Top 15 High-Quality Consensus Patterns",
      digits = 3)
```

### Interpretation

High-quality patterns represent strong candidates for functional stalling peptides based on:

- **Large cluster size**: Many independent observations of the pattern
- **High conservation**: Sequence preservation suggests functional importance
- **Known motif types**: Match to characterized stalling motifs (RAGP, QAPP, etc.)

---

# Validation Results

## Known Peptide Recovery

```{r validation-summary}
if (nrow(validation) > 0) {
  # Best hit for each peptide
  best_hits <- validation %>%
    group_by(peptide_id) %>%
    slice_min(evalue, n = 1) %>%
    ungroup()

  # Recovery by motif type
  if ("motif_type" %in% colnames(best_hits)) {
    recovery_by_motif <- best_hits %>%
      count(motif_type, name = "count")

    kable(recovery_by_motif,
          col.names = c("Motif Type", "Peptides Recovered"),
          caption = "Recovery of Known Stalling Peptides by Motif Type")
  }

  # Top hits
  top_hits <- best_hits %>%
    arrange(evalue) %>%
    head(10) %>%
    select(peptide_id, cluster_id, evalue, score, organism)

  kable(top_hits,
        col.names = c("Peptide", "Cluster", "E-value", "Score", "Organism"),
        caption = "Top 10 Validation Hits",
        digits = c(0, 0, 2, 1, 0))

} else {
  cat("No validation results available.\n")
}
```

## Recovery Rate

```{r recovery-rate}
if (nrow(validation) > 0 && comparison$known_peptides_recovered > 0) {
  recovery_text <- sprintf(
    "**Recovery rate**: %d known stalling peptides were successfully identified by the discovered cluster HMMs.",
    comparison$known_peptides_recovered
  )
  cat(recovery_text)
} else {
  cat("Validation data not available or no peptides recovered.")
}
```

### Notable Recoveries

```{r notable-recoveries}
if (nrow(validation) > 0) {
  # Highlight key stalling peptides
  key_peptides <- c("SecM", "TnaC", "MifM", "CydA")

  notable <- validation %>%
    filter(str_detect(peptide_id, paste(key_peptides, collapse="|"))) %>%
    group_by(peptide_id) %>%
    slice_min(evalue, n = 1) %>%
    ungroup() %>%
    select(peptide_id, cluster_id, evalue, score)

  if (nrow(notable) > 0) {
    kable(notable,
          col.names = c("Peptide", "Cluster", "E-value", "Score"),
          caption = "Recovery of Key Known Stalling Peptides",
          digits = c(0, 0, 2, 1))
  }
}
```

---

# Sequence Logos

## Top Cluster Logos

*Note: Sequence logos will be generated in issue #12 using the Skylign API.*

Placeholder logos have been created for the top 20 clusters. These will be replaced with publication-quality sequence logos showing:

- Position-specific amino acid frequencies
- Conservation information content
- Key motif features (RAGP, QAPP, etc.)

---

# Candidate Stalling Peptides

## High-Confidence Candidates

Based on cluster size, conservation, and motif patterns, the following are high-confidence candidates for experimental validation:

```{r candidates}
candidates <- consensus %>%
  filter(quality_score > 0.7) %>%
  arrange(desc(quality_score)) %>%
  head(10) %>%
  mutate(
    recommendation = case_when(
      motif_type %in% c("RAGP", "RAPG") ~ "High priority - SecM-like",
      motif_type == "QAPP" ~ "High priority - B. subtilis-like",
      motif_type == "GP" & quality_score > 0.8 ~ "Medium priority - novel pattern",
      TRUE ~ "Medium priority"
    )
  ) %>%
  select(cluster_id, motif_type, cluster_size, mean_conservation,
         consensus_sequence, recommendation)

kable(candidates,
      col.names = c("Cluster", "Motif", "Size", "Conservation",
                    "Consensus", "Recommendation"),
      caption = "Top 10 Candidate Stalling Peptides for Experimental Validation",
      digits = 3)
```

## Novel Patterns

Clusters with high conservation but not matching known motif types represent potentially novel stalling peptides:

```{r novel-patterns}
novel <- consensus %>%
  filter(motif_type == "GP", quality_score > 0.6) %>%
  arrange(desc(quality_score)) %>%
  head(5) %>%
  select(cluster_id, cluster_size, mean_conservation, quality_score, consensus_sequence)

if (nrow(novel) > 0) {
  kable(novel,
        col.names = c("Cluster", "Size", "Conservation", "Quality", "Consensus"),
        caption = "Novel GP-Containing Patterns",
        digits = 3)
} else {
  cat("No novel patterns identified (all match known motif types).\n")
}
```

---

# Methods

## Databases Searched

- **UniProt Bacteria**: Curated bacterial proteins
- **Pfam**: Protein domain annotations (v. current)

## Computational Pipeline

### Approach 1: Seed-Based Discovery

1. Parse known stalling peptides from literature (n=47)
2. Split by motif type (RAGP, QAPP, etc.)
3. Build seed HMMs with `hmmbuild`
4. Search bacterial proteomes with `hmmsearch`
5. Iterative refinement

### Approach 2: Unbiased GP Discovery

1. **GP Extraction**: Identify all GP-containing sequences with context (30 upstream, 15 downstream)
2. **Domain Annotation**: Run `hmmscan` against Pfam database
3. **Inter-domain Filtering**: Classify GP motifs as inter-domain vs intra-domain
4. **Clustering**: Group similar GP contexts using CD-HIT/MMseqs2
   - Identity threshold: `r ifelse(exists("config"), config$prokaryotic$clustering_identity, "70%")`
   - Coverage threshold: `r ifelse(exists("config"), config$prokaryotic$clustering_coverage, "80%")`
5. **Conservation Analysis**: Align each cluster with MUSCLE, calculate Shannon entropy
6. **Consensus Identification**: Extract consensus sequences, build HMMs
7. **Validation**: Search against known stalling peptides

## Parameters

- **E-value threshold** (domain annotation): 1e-3
- **Pfam cutoff**: Gathering thresholds
- **Minimum cluster size**: `r ifelse(exists("config"), config$prokaryotic$min_cluster_size, "5")`
- **Minimum conservation**: `r ifelse(exists("config"), config$prokaryotic$min_conservation, "60%")`

## Software

- **HMMER** 3.3+: Profile HMM searches
- **MUSCLE**: Multiple sequence alignment
- **Easel** tools: Format conversion
- **Biopython**: Sequence processing
- **R/tidyverse**: Analysis and reporting

---

# Conclusions

## Summary

This analysis successfully identified **`r comparison$n_clusters` distinct GP motif clusters** from prokaryotic proteomes, with **`r nrow(consensus)` high-quality consensus patterns** showing significant conservation.

**Key achievements:**

1. ✅ Comprehensive GP motif extraction from bacterial proteins
2. ✅ Domain-boundary enrichment analysis
3. ✅ Pattern clustering and conservation analysis
4. ✅ Recovery of known stalling peptides: **`r comparison$known_peptides_recovered`**
5. ✅ Identification of novel candidate patterns

## Biological Insights

The **`r sprintf("%.1f%%", comparison$interdomain_fraction * 100)`** enrichment of GP motifs in inter-domain regions supports the model that:

- Stalling peptides preferentially occur at domain boundaries
- Co-translational folding dynamics influence stalling site evolution
- Multi-domain proteins may use stalling for regulated assembly

## Experimental Recommendations

**High-priority targets for validation:**

```{r final-recommendations}
priority_targets <- consensus %>%
  filter(quality_score > 0.7) %>%
  arrange(desc(quality_score)) %>%
  head(5) %>%
  pull(cluster_id)

cat(sprintf("- Clusters: %s\n", paste(priority_targets, collapse=", ")))
cat(sprintf("- Motif types represented: %s\n",
            paste(unique(consensus %>%
                          filter(cluster_id %in% priority_targets) %>%
                          pull(motif_type)), collapse=", ")))
```

**Validation approach:**

1. Synthesize reporter constructs with candidate peptides
2. In vitro translation assays (ribosome profiling)
3. In vivo reporter assays (GFP fusion)
4. Toe-printing or cryo-EM for stalling confirmation

---

# Session Info

```{r session-info}
sessionInfo()
```

---

**Report generated**: `r Sys.time()`
**Pipeline version**: 2A-peptide-search v1.0
**Contact**: See repository README for details
