---
title: "GP Motif Analysis"
subtitle: "Comprehensive analysis of Gly-Pro motifs in prokaryotic proteins"
date: today
---

## Overview

The Gly-Pro (GP) dipeptide is central to ribosomal stalling. This analysis examines GP motif distribution, context, and conservation across prokaryotic proteomes.

## GP Motif Extraction

### Parameters

```yaml
prokaryotic:
  gp_upstream: 30      # Residues upstream of GP
  gp_downstream: 15    # Residues downstream of GP
```

### Expected Distribution

GP motifs occur throughout proteins but functional stalling sites have distinctive features:

| Feature | Random GP | Stalling GP |
|---------|-----------|-------------|
| Position | Any | Inter-domain |
| Upstream | Variable | Conserved pattern |
| Downstream | Variable | Often P or charged |

## Context Analysis

### Upstream Patterns

Functional stalling peptides show conserved patterns upstream of GP:

```
-30    -20    -10    GP
 |      |      |     ||
 L---F--W------R/K--AGP  (SecM-like)
 W------E------G----GP   (TnaC-like)
```

### Position-Specific Conservation

| Position | SecM-like | TnaC-like | Random |
|----------|-----------|-----------|--------|
| -1 (A/G) | High | Medium | Low |
| -2 (R/K) | High | Low | Low |
| -3 | Variable | Variable | Variable |
| -10 (W) | High | High | Low |

## Domain Context

### Inter-Domain vs Intra-Domain

GP motifs are classified by their domain context:

```{mermaid}
%%| label: fig-domain-context
%%| fig-cap: "GP motif domain contexts"
flowchart LR
    subgraph Protein
        D1[Domain A] --> L[Linker + GP] --> D2[Domain B]
    end

    subgraph Types
        T1[Inter-domain GP]
        T2[Intra-domain GP]
        T3[Terminal GP]
    end
```

- **Inter-domain**: Between two annotated Pfam domains
- **Intra-domain**: Within a single domain
- **Terminal**: Near N- or C-terminus

### Enrichment Analysis

Functional stalling GPs are enriched in inter-domain regions:

| Context | All GP | Stalling GP | Enrichment |
|---------|--------|-------------|------------|
| Inter-domain | 15% | 60% | 4x |
| Intra-domain | 70% | 25% | 0.4x |
| Terminal | 15% | 15% | 1x |

## Clustering Results

GP motif sequences were clustered using MMseqs2 at 70% sequence identity to identify conserved families.

```{r}
#| label: setup-clustering
#| code-fold: true
#| message: false
#| warning: false

library(tidyverse)
library(scales)

cluster_file <- here::here("results/prokaryotic/gp_analysis/bacteria/gp_clusters.tsv.gz")

if (file.exists(cluster_file)) {
  clusters_df <- read_tsv(cluster_file, show_col_types = FALSE)
  data_available <- TRUE
} else {
  data_available <- FALSE
  message("Cluster data not found at ", cluster_file)
}
```

### Summary Statistics

```{r}
#| label: cluster-summary
#| code-fold: true

if (data_available) {
  n_total_motifs <- nrow(clusters_df)
  n_unique_seqs <- n_distinct(clusters_df$context_sequence)
  n_clusters <- n_distinct(clusters_df$cluster_id)
  duplication_ratio <- n_total_motifs / n_unique_seqs

  cluster_stats <- clusters_df |>
    group_by(cluster_id) |>
    summarise(
      total_motifs = n(),
      unique_seqs = n_distinct(context_sequence),
      .groups = "drop"
    )

  n_singletons <- sum(cluster_stats$unique_seqs == 1)
  pct_singletons <- 100 * n_singletons / n_clusters

  tibble(
    Metric = c(
      "Total GP motifs",
      "Unique sequences",
      "Total clusters",
      "Singleton clusters",
      "Duplication ratio",
      "Largest cluster (unique)",
      "Largest cluster (total)"
    ),
    Value = c(
      comma(n_total_motifs),
      comma(n_unique_seqs),
      comma(n_clusters),
      paste0(comma(n_singletons), " (", round(pct_singletons, 1), "%)"),
      paste0(round(duplication_ratio, 2), "x"),
      comma(max(cluster_stats$unique_seqs)),
      comma(max(cluster_stats$total_motifs))
    )
  ) |>
    knitr::kable()
}
```

### Cluster Size Distribution

```{r}
#| label: fig-cluster-size-dist
#| fig-cap: "Distribution of cluster sizes by unique sequences and total motifs"
#| code-fold: true
#| fig-width: 10
#| fig-height: 4

if (data_available) {
  plot_data <- cluster_stats |>
    pivot_longer(
      cols = c(unique_seqs, total_motifs),
      names_to = "count_type",
      values_to = "count"
    ) |>
    mutate(count_type = recode(count_type,
      unique_seqs = "Unique Sequences",
      total_motifs = "Total Motifs"
    ))

  ggplot(plot_data, aes(x = count, fill = count_type)) +
    geom_histogram(bins = 50, alpha = 0.7) +
    scale_x_log10(labels = comma) +
    scale_y_log10(labels = comma) +
    facet_wrap(~count_type, ncol = 2) +
    labs(
      x = "Cluster Size",
      y = "Number of Clusters (log scale)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

### Top Clusters

```{r}
#| label: top-clusters-table
#| code-fold: true

if (data_available) {
  cluster_stats |>
    slice_max(unique_seqs, n = 15) |>
    mutate(duplication = round(total_motifs / unique_seqs, 2)) |>
    rename(
      `Cluster ID` = cluster_id,
      `Unique Seqs` = unique_seqs,
      `Total Motifs` = total_motifs,
      Duplication = duplication
    ) |>
    knitr::kable()
}
```

### Duplication Analysis

The relationship between unique sequences and total motifs reveals how duplicated each cluster is.

```{r}
#| label: fig-duplication-scatter
#| fig-cap: "Unique sequences vs total motifs per cluster. Points above the diagonal have duplicated sequences."
#| code-fold: true
#| fig-width: 6
#| fig-height: 6

if (data_available) {
  multi_clusters <- cluster_stats |>
    filter(unique_seqs > 1)

  ggplot(multi_clusters, aes(x = unique_seqs, y = total_motifs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(alpha = 0.3, size = 1) +
    scale_x_log10(labels = comma) +
    scale_y_log10(labels = comma) +
    labs(
      x = "Unique Sequences per Cluster",
      y = "Total Motifs per Cluster"
    ) +
    theme_minimal()
}
```

### Cluster Size Breakdown

```{r}
#| label: cluster-size-breakdown
#| code-fold: true

if (data_available) {
  cluster_stats |>
    mutate(size_category = case_when(
      unique_seqs == 1 ~ "1 (singleton)",
      unique_seqs <= 5 ~ "2-5",
      unique_seqs <= 10 ~ "6-10",
      unique_seqs <= 50 ~ "11-50",
      unique_seqs <= 100 ~ "51-100",
      TRUE ~ ">100"
    )) |>
    mutate(size_category = factor(size_category,
      levels = c("1 (singleton)", "2-5", "6-10", "11-50", "51-100", ">100")
    )) |>
    group_by(size_category) |>
    summarise(
      Clusters = n(),
      `Unique Seqs` = sum(unique_seqs),
      `Total Motifs` = sum(total_motifs),
      .groups = "drop"
    ) |>
    mutate(`% Clusters` = round(100 * Clusters / n_clusters, 1)) |>
    rename(`Size Category` = size_category) |>
    knitr::kable()
}
```

## Downstream Analysis

### Codon Usage

Stalling peptides may show:

- Rare codon usage upstream of GP
- Specific codon at P position
- mRNA secondary structure

### Ribosome Profiling Validation

Compare GP positions to ribosome profiling data:

- Ribosome occupancy peaks
- Pause site mapping
- Translation rate analysis

## Output Files

```
results/prokaryotic/gp_analysis/
├── {database}/
│   ├── all_gp_motifs.tsv.gz      # All extracted GP contexts
│   ├── all_gp_sequences.fasta.gz # Full sequences
│   ├── domain_annotations.tsv    # Pfam domain hits
│   ├── interdomain_gp.tsv        # Filtered inter-domain
│   └── clusters/
│       ├── cluster_assignments.tsv
│       └── cluster_{N}.fasta
```

## Quality Metrics

### Filtering Statistics

| Filter | Sequences Removed |
|--------|-------------------|
| No GP motif | N/A (pre-filtered) |
| Too short (<45 aa) | TBD |
| Intra-domain | TBD |
| Low-complexity | TBD |

### Cluster Quality

| Metric | Threshold | Purpose |
|--------|-----------|---------|
| Cluster size | ≥5 sequences | Statistical power |
| Conservation | ≥0.6 | Functional signal |
| Identity | 70% clustering | Reduce redundancy |
